Computing likelihoods under the SMC for a general class of ARGs.

### 2023-11-13 paper.tex
start introduction
update intro and discussion

\section{Abstract}
%% WHAT IS THE MAIN FOCUS OF THE PAPER?

@@

% Main motivation:
Major progress in ARG inference.

Result is whole range of inference methods, all inferring an ARG with their own specific 
characteristics.
Many of these scalable heuristic methods do not explicitly infer recombination events.

%ARGs that are inferred here lack information to enable us to compute 
% a likelihood under the Hudson coalescent.
% Missing: exact time to recombination + non-inferrable recombinations

Currently no way to connect the output of the diverse range of ARG inference methods back 
to a model that would allow us to perform a probabilistic exploration of the 
ARG space around the point estimate, (or likelihood-based inference, ...)

% What are ARGs, their relevance and importance

% associating likelihoods with ARGs

% previous likelihood calculations on ARGs Kuhner and Felsenstein (2000), ARGweaver

% history of approximating the coalescent: 
%SMC (McVean and Cardin, 2005) SMC' (Marjoram and Wall, 2006).
%Also heuristic approximations: Li and Stephens (2003) extending Stephens and Donnelly (2000)
%These have enabled (scalable) ARG inference methods: ARGweaver (relying on SMC, limited 
%to smaller samples, but ability to sample from posterior, and full ARG), and Relate, 
%tsinfer/tsdate (copying algorithm), ArgNeedle (using SMC) 
%Clear distinction into two groups: ARGWeaver infers actual ARG.
% what does kwarg do?
%Whereas the other approaches are heuristic in nature. For example, where ARGweaver has 
%the built in assumption of neighbouring trees being separated by single Subtree Prune 
%Regraft (SPR) (Song citation?), this is not the case for any of the heuristic data 
%structures. (other peculiarities: ...).

% computing likelihoods on ARGs; from Kuhner and Felsenstein to ARGWeaver

% SMC and relatives are useful: lot of focus however on integrating over the distribution of trees to get Ne(t). \cite{PSMC, ...}
% Here the idea is to use the SMC in a different way that resonates more with the true direction of flow
% of genetic/haplotype information.

% results in a concrete way to compute likelihoods for *any* inferred ARG (integrating out the recombination time)
% meaning: e.g. no requirements on adjacent trees only differing by a single SPR, (other?)

% In this paper, we attempt to address these problems by
% providing a precise definition of an ARG that is general enough to encompass
% the output of popular simulation and inference methods.

% ADD: remark on fact that we are only dealing with topology and branch length information
% adding likelihood of observing particular mutation configuration given an ARG is straightforward.

@@

We can now define $C_i(t) = \{X_j \in L(t) | I_{ij}(t) = 1\}$ such that $|C_i(t)| = 
I_{i}(t) = \sum_{j} I_{ij}(t)$.
Because $I_{ij}$ is not reciprocal, each extant lineage has their own unique set of 
lineages $C_i(t)$ it can coalesce with. This will simplify the likelihood computations 
(see \ref{par:liks). Note that although recombination affects $C_i(t)$ it does not 
affect $I_{i}$.\\

@@

\section{Discussion}
% summary: what did we do here
Here, we have introduced a scalable algorithm to compute likelihoods 
under the SMC for a general class of ARGs.

Impact both on ways of doing ARG inference itself and quantifying the uncertainty associated 
with that as well as performing statistical inference of past demography.

* statistical inference *
This is an important step towards bridging the gap between the tremendous 
recent advances in ARG inference and the current state of the art statistical inference of 
past demography (but see ). 
% some/many of these approaches only use the information present in marginal trees.
The provided theory and associated algorithm will function as a building block of a 
demography-aware genealogy sampler. Current approach couple of major advantages:
* any type of inferred ARG as a starting point. Potentially implying a massive reduction in 
MCMC runtime by reducing the necessary burnin period.
* by integrating out the exact time to the recombination event we further restrict the size 
of ARG space that needs to be explored.
* exploit the efficiencies of the tree sequence data structure (see \citep{Mahmoudi_bayesian_2022})
* we can readily incorporate $N_e$-changes as well as multiple populations and migration.

* ARG inference *
Combining ARG inference methods: tsinfer for first few generations (large sample sizes), where 
the initial phase is biased relative to the Wright-Fisher model when the sample size is large 
\citep{bhaskar_distortion_2014}, or when the considered events are very recent \citep{wakeley_gene_2012}
% tsinfer not model-based, minimal assumptions, heuristic, scalable

quantify uncertainty
* polytomies represent uncertainty:
*  systematicallty break them by sampling from the coalescent rather then randomly
* associate uncertainty as metadata to parent node representing uncertainty in the
* interval spanned by the polytomy
* can something similar be done for stacked recombinations?

>>>>

\section{Abstract}
This means there currently is no way to connect the output of the diverse 
range of ARG inference methods back 
to a model that would allow us to perform a probabilistic exploration of the 
ARG space around the point estimate, or likelihood-based inference

@@

The underlying genealogical history that relates a sample of individuals back in time
contains all possible information this sample can provide on the history of demography, 
selection and recombination for the population of interest. 
Reconstructing these genealogies with certainty is impossible however. 
As such, in the past the data structure capturing these inheritance path 
or Ancestral Recombination Graph (ARG), has usually been considered a nuisance 
variable to be integrated out during statistical inference 
\citep{kuhner_2000, griffiths_ancestral_1996}.

% explicitly outputs an ARG
ARGweaver, generates these ARGs explicitly again using MCMC sampling, by discretizing time 
they can apply HMM-based algorithms to efficiently generate new samples from the distribution of 
possible ARGs.

% model-based approaches vs

% recent advancements: heuristics
Recently, major advancements 
relying on heuristic rather than model-based approaches
made ARG inference scalable.
\citep{rasmussen_2014, kelleher_2019, speidel_2019, schaefer_2021, wohns_2022, 
zhang_biobank-scale_2023, zhan_towards_2023}
have raised the realistic prospect of ARG-based analysis becoming a standard 
part of the population and statistical genetics toolkit
This has also meant however that
we are now using the term ARG for a varying range of approxiate structures
as generated by these inference methods \citep{wong_a-general_2023}.

Firstly, in contrast to model-based inference methods like \argweaver, 
these models only deal with recombination in an implicit sense.

For example, where ARGweaver has 
the built in assumption of neighbouring trees being separated by single Subtree Prune 
Regraft (SPR) [REF Song et al.], this is not the case for any of the heuristic data 
structures.
Lack recombination nodes in the ARG, crucial in ability to compute likelihoods 
under Hudson. \citep{Wong_a-general_2023}

Secondly, the lack of an underlying model also means that
we have no way to deal with uncertainty, an issue that is 
resolved differently by each inference tool.

Besides not having a unified output format, The degree of completeness 
in which genetic inheritance from ancestors to 
descendants is documented by each of these methods varies extensively.

This means there currently is no way to connect the output of the diverse 
range of ARG inference methods back 
to a model that would allow us to perform a probabilistic exploration of the 
ARG space around the point estimate, or likelihood-based inference 
(that does not only consider individual marginal trees).

[TO ADD] mention the SMC \citep{mcvean_approximating_2004}.
SMC and relatives are useful: lot of focus however on integrating over 
the distribution of trees to get Ne(t) [ADD REF: PSMC, MSMC]
Here the idea is to use the SMC in a different way that resonates more 
with the true direction of flow of genetic/haplotype information.
Results in a concrete way to compute likelihoods for *any* inferred ARG 
by integrating out the time to recombination.
(integrating out the recombination time)
and formulate an algorithm 
that is general in the sense that it can deal with the various levels of 
completeness that characterise the fields currents attempts at reconstucting genome-wide 
genealogies.

[TO ADD] remark on fact that we are only dealing with topology and branch length information
adding likelihood of observing particular mutation configuration given an ARG is straightforward.)

@@

We can now define $C_i(t) = \{X_j \in L(t) | I_{ij}(t) = 1\}$ such that $|C_i(t)| = 
I_{i}(t) = \sum_{j} I_{ij}(t)$.
Because $I_{ij}$ is not reciprocal, each extant lineage has their own unique set of 
lineages $C_i(t)$ it can coalesce with. This will simplify the likelihood computations 
(see \ref{par:liks}). Note that although recombination affects $C_i(t)$ it does not 
affect $I_{i}$.\\

@@

\section{Discussion}
% summary: what did we do here
Here, we have introduced a scalable algorithm to compute likelihoods 
under the SMC for a general class of ARGs. This work will contribute significantly 
to ARG-based analysis becoming a standard part of the population 
and statistical genetics toolkit. 
In particular, we see three main applications for the work presented here. 
Firstly, the ability to associate inferred ARGs with a likelihood given a demographic 
model and a set of parameters will help bridge the gap between the tremendous recent 
advances in ARG inference and the current state of the art statistical inference of 
past demography. Secondly, hybrid ARG-inference methods combining the best of both 
heuristic and model based inference approaches. Thirdly, quantify uncertainty.

1/ statistical inference
This is an important step towards bridging the gap between the tremendous 
recent advances in ARG inference and the current state of the art statistical inference of 
past demography (but see ): currently most/all methods perform ARG-based analysis 
by extracting mariginal trees from the ARG \citep{hejase_2022} % likelihood-based demographic inference on local trees (ref?) 
The provided theory and associated algorithm will function as a building block of a 
demography-aware genealogy sampler. Current approach couple of major advantages:
* any type of inferred ARG as a starting point. Potentially implying a massive reduction in 
MCMC runtime by reducing the necessary burnin period.
* by integrating out the exact time to the recombination event we further restrict the size 
of ARG space that needs to be explored.
* exploit the efficiencies of the tree sequence data structure (see \citep{Mahmoudi_bayesian_2022})
* we can readily incorporate $N_e$-changes as well as multiple populations and migration.\\

Biobank scale datasets, first few generations violate coalescent assumptions. Here we use heuristic 
method. Once number of lineages has reduced sufficiently, model-based approach can be used. One 
of the advantages of formulating this backwards in time. 
Combining ARG inference methods: tsinfer for first few generations (large sample sizes), where 
the initial phase is biased relative to the Wright-Fisher model when the sample size is large 
\citep{bhaskar_distortion_2014}, or when the considered events are very recent \citep{wakeley_gene_2012}\\

Both of these future directions tie in with the ability to quantify uncertainty by sampling 
from an underlying distribution. This will in particular enable us to deal with those 
parts in both time and along the genome for which we have insufficient information to confidently 
reconstruct the ARG. For example the presence of polytomies in \tsinfer represents 
uncertainty. Systematicallty break them by sampling from the coalescent rather then randomly.
Quantifying uncertainty would not necessarily always have to be associated with dealing with 
the entire subset of all compatible genealogies. Uncertainty could also be quantified and passed 
on as metadata to a parent node representing the uncertainty in the interval spanned by the 
polytomy.

% can something similar be done for stacked recombinations?

[TO DO] Comment on accuracy. The described algorithm has been implemented in such a way that 
any valid tree sequence will result in a likelihood value. This also holds for any tree sequence 
that does not meet the assumptions. We see it as future work to determine to what extent 
particular deviations from these assumptions will affect the shape of the likelihood surface given 
a set of parameters.

% final conclusion
